BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-06-27 background.tex
Structure revised

\chapter{API Hooking: State of The Art}
In literature there are many different implementations of API hooking. The objective of this chapter is to provide an outline of the various approaches utilized to hook functions in DLLs, outlining the benefits and the limitations of each technique, with a strong focus on their detection by malicious software. In particular, the focus will be on user space API hooking of \texttt{Win32} binaries, since this is BlueTracer's current field of application. Obviously, as it is the norm in malware analysis, it also assumed that the program under study is only available in binary form.

Depending on their underlying implementation, API hooking techniques can be divided in three broad categories: \textbf{Binary Rewriting Based}, \textbf{Virtul Machine Introspection (VMI) Based} and \textbf{Dynamic Binary Instrumentation (DBI) based}.

\iffalse
backmatter turns off chapter numbering
\fi
\begin{section}{Binary Rewriting Based Hooking}

@@

\subsection{Import Address Table (IAT) patching}

@@

\subsection{Export Address Table (EAT) patching}

@@

\subsection{Inline hooking}

@@

On the other hand, a debugger is much easier to detect. In fact, there exist specific Windows APIs whose purpose is to find out whether or not the current process is being debugged. For example, \texttt{IsDebuggerPresent} allows to determine if the calling process is running under a debugger, while \texttt{CheckRemoteDebuggerPresent} checks for the presence of a debugger in a separate process. 
In addition, the \texttt{INT 3} instruction in an API entry point immediately gives away the debugger's presence \cite{HookingDetection}.
\end{section}

\begin{section}{Virtual Machine Introspection (VMI) \\ 
Based Hooking}

@@

\end{section}

\begin{section}{Dynamic Binary Instrumentation (DBI) \\
Based Hooking}

Dynamic Binary Instrumentation (DBI) is an analysis technique in which the behavior of a binary application is inspected at run-time via the injection of instrumentation code. Such code, after being injected, executes as a component of the ordinary instruction flow, allowing to learn information about the behavior and the state of a sample at different points during its execution \cite{DBI}. In DBI based hooking, the learnt information refers to which APIs are called and, possibly, with which arguments and return values.

@@

\end{section}

\begin{section}{Conclusion}

In this chapter it was shown how the state of the art API hooking techniques suffer from a number of remarkable shortcomings, especially when dealing with evasive malwares. In fact, binary rewriting based hooking techniques are all easily detectable, while VMI, although harder to uncover, is affected by the \textit{semantic gap problem}. Finally, existing DBI based API tracing tools are not accompanied by adequate cloaking mechanisms and are limited in the amount of logged information. The aforementioned issues indicate that there is a need for a robust API tracer, specialized in the analysis of evasive malwares and with extensive logging capabilites. This is the rationale at the heart of BlueTracer.

\end{section}

>>>>

\chapter{Background}

In literature there are many different implementations of API hooking. The objective of this chapter is to provide an outline of the various approaches utilized to hook functions in DLLs, outlining the benefits and the limitations of each technique, with a strong focus on their detection by malicious software. In particular, the focus will be on user space API hooking of \texttt{Win32} binaries, since this is BlueTracer's current field of application. Obviously, as it is the norm in malware analysis, it also assumed that the program under study is only available in binary form.

Depending on their underlying implementation, API hooking techniques can be divided in three broad categories: \textbf{Binary Rewriting} based, \textbf{Virtul Machine Introspection (VMI)} based and \textbf{Dynamic Binary Instrumentation (DBI)} based.

\section{Binary Rewriting}

@@

\subsection{Import Address Table Patching}

@@

\subsection{Export Address Table Patching}

@@

\subsection{Inline Hooking}

@@

On the other hand, a debugger is much easier to detect. In fact, there exist specific Windows APIs whose purpose is to find out whether or not the current process is being debugged. For example, \texttt{IsDebuggerPresent} allows to determine if the calling process is running under a debugger, while \texttt{CheckRemoteDebuggerPresent} checks for the presence of a debugger in a separate process. 
In addition, the \texttt{INT 3} instruction in an API entry point immediately gives away the debugger's presence \cite{HookingDetection}.

\section{Virtual Machine Introspection}

@@

\section{Dynamic Binary Instrumentation}

Dynamic Binary Instrumentation (DBI) is an analysis technique in which the behavior of a binary application is inspected at run-time via the injection of instrumentation code. Such code, after being injected, executes as a component of the ordinary instruction flow, allowing to learn information about the behavior and the state of a sample at different points during its execution \cite{DBI}. We will further elaborate on DBI throughout\mynote{add ref} Chapter~\ref{...}.

In DBI based hooking, the learnt information refers to which APIs are called and, possibly, with which arguments and return values.

@@

\section{Conclusion}

In this chapter we showed how the state of the art API hooking techniques suffer from a number of remarkable shortcomings, especially when dealing with evasive malware. In fact, binary rewriting based hooking techniques are all easily detectable, while VMI, although harder to uncover, is affected by the \textit{semantic gap problem}. Finally, existing DBI-based API tracing tools are not accompanied by adequate cloaking mechanisms and are limited in the amount of logged information. The aforementioned issues indicate that there is a need for a robust API tracer, specialized in the analysis of evasive malware and with extensive logging capabilities. This is the rationale at the heart of BlueTracer.

### 2018-06-27 implementation.tex
Structure revised

\chapter{Implementation}
In this chapter BlueTracer's implementation will be discussed, providing a detailed account on how the tool is organized, on the design choices which were made during its development process, on how encountered challenges were dealt with and on the decisions which were taken to improve performance.

BlueTracer, being a part of Blue Pill, is also implemented using Pin, a dynamic binary instrumentation framework by Intel, which is vastly utilized for program analysis, testing of software and in the security field. The version of Pin used for the development of the tool is 3.5, in order to benefit from the notable improvements, both in terms of execution speed and offered features, which were introduced going from the 2.14 release to the 3.x series. Pin comes with its own OS-agnostic and compiler-agnostic runtime, called PinCRT. PinCRT exposes three layers: a generic operating system interface providing basic OS services (e.g. process and thread control), together with C and C++03 (without RTTI) runtime layers, for writing instrumentation and analysis routines \cite{Pin}.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.8]{Figures/BlueTracer.pdf}
\caption{\textit{BlueTracer's high level structure}}
\end{figure}

@@

In Pin, the Native API is intercepted right before \texttt{sysenter} is executed. This could be inferred by the fact that, in \texttt{TraceSysCallEntry}, the instruction pointer (obtainable via the \texttt{PIN\_GetContextReg} API) contains \texttt{sysenter}'s address. In light of this, in addition to the fact that each Native API calls \texttt{KiFastSystemCall} without setting a stack frame, the application's stack during the execution of \texttt{TraceSysCallEntry} is in the following state (\textit{Figure 3.3}).
\\
\begin{figure}[h]
\centering
\includegraphics[scale=1]{Figures/StackMainImage.pdf}
\caption{\textit{Stack before} \texttt{sysenter} \textit{execution}}
\end{figure}

>>>>

\chapter{Architecture and Implementation}
In this chapter we discuss the implementation of BlueTracer, providing a detailed account on how the tool is organized, on the design choices which were made during its development process, on how encountered challenges were dealt with and on the decisions which were taken to improve performance.

\section{Overview}

BlueTracer\mynote{This section requires significant extensions}, being a part of Blue Pill, is also implemented using Pin, a dynamic binary instrumentation framework by Intel, which is vastly utilized for program analysis, testing of software and in the security field. The version of Pin used for the development of the tool is 3.5, in order to benefit from the notable improvements, both in terms of execution speed and offered features, which were introduced going from the 2.14 release to the 3.x series. Pin comes with its own OS-agnostic and compiler-agnostic runtime, called PinCRT. PinCRT exposes three layers: a generic operating system interface providing basic OS services (e.g. process and thread control), together with C and C++03 (without RTTI) runtime layers, for writing instrumentation and analysis routines \cite{Pin}.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{Figures/BlueTracer.pdf}
\caption{\textit{BlueTracer's high level structure}}
\end{figure}

@@

In Pin, the Native API is intercepted right before \texttt{sysenter} is executed. This could be inferred by the fact that, in \texttt{TraceSysCallEntry}, the instruction pointer (obtainable via the \texttt{PIN\_GetContextReg} API) contains \texttt{sysenter}'s address. In light of this, in addition to the fact that each Native API calls \texttt{KiFastSystemCall} without setting a stack frame, the application's stack during the execution of \texttt{TraceSysCallEntry} is in the following state (\textit{Figure 3.3}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{Figures/StackMainImage.pdf}
\caption{\textit{Stack before} \texttt{sysenter} \textit{execution}}
\end{figure}

### 2018-06-27 introduction.tex
Structure revised

The goal of this thesis is to present \textbf{BlueTracer}, a robust library and system call tracer for Windows applications, specialized in the monitoring of evasive malwares. BlueTracer is based on the Intel Pin \cite{Pin} dynamic binary instrumentation (DBI) framework and is able to counteract malwares' anti-evasion measures thanks to its integration with BluePill, a software toolkit built on top of a DBI layer which allows the simulation of the execution environment a particular malware was designed for and conceals any virtualization artifacts and setup details which might set off evasion \cite{BluePill}. BlueTracer is capable of tracing the input values, the output values and the return values of an extremely wide range of system calls (including Windows Native APIs) and API calls. Moreover, it also supports the tracing of Windows callbacks functions and Windows asynchronous procedure calls (APC). The tool was tested on a benign application aimed at assessing how good an anti-malware system is against evasion techniques and on actual evasive malwares, proving to be effective in both tracing the samples' activity and remaining undetected.

>>>>

\paragraph{Contributions.}
Throghout this thesis we present \textbf{BlueTracer}, a robust library and system call tracer for Windows applications, specialized in the monitoring of evasive malwares. BlueTracer is based on the Intel Pin \cite{Pin} dynamic binary instrumentation (DBI) framework and is able to counteract malwares' anti-evasion measures thanks to its integration with BluePill, a software toolkit built on top of a DBI layer which allows the simulation of the execution environment a particular malware was designed for and conceals any virtualization artifacts and setup details which might set off evasion \cite{BluePill}. BlueTracer is capable of tracing the input values, the output values and the return values of an extremely wide range of system calls (including Windows Native APIs) and API calls. Moreover, it also supports the tracing of Windows callbacks functions and Windows asynchronous procedure calls (APC). The tool was tested on a benign application aimed at assessing how good an anti-malware system is against evasion techniques and on actual evasive malwares, proving to be effective in both tracing the samples' activity and remaining undetected.
